---
title: "Critique of model evaluation by WA Dept of Ecology"
author: "Mark Scheuerell"
date: "10/31/2019"
output:
  pdf_document:
    highlight: haddock
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "pdf")

if(!require("captioner")) {
  devtools::install_github("adletaw/captioner")
  library("captioner")
}

## set default caption delimter
fig_cap <- captioner(suffix = ".")
## initialize figure number
cnt <- 1
```


# Error in assumptions

In the section titled "Uncertainty in Dissolved Oxygen Depletion Estimates" (p59), there is an error in the assumed relationship between the standard deviation of the predictions and the root mean squared error (RMSE) of the predictions. Specifically, the document incorrectly suggests that

$$
\text{Var}(\hat{y}) = \text{RMSE}(\hat{y})^2 ~ \Rightarrow ~ \text{SD}(\hat{y}) = \text{RMSE}(\hat{y}).
$$

Specifically, the standard deviation of ${\hat{y}}$ is based upon differences between the predicted data points $\hat{y}_i$ and their mean $m_{\hat{y}}$

$$
\text{SD}(\hat{y}) = \sqrt{\frac{(\hat{y}_i - m_{\hat{y}})^2}{N}},
$$

whereas the RMSE of $\hat{y}$ is based upon differences between the predicted data points and their corresponding observed values $x_i$

$$
\text{RMSE}(\hat{y}) = \sqrt{\frac{(\hat{y}_i - y_i)^2}{N}}.
$$

Thus,

$$
\begin{aligned}
\sqrt{\frac{(\hat{y}_i - m_{\hat{y}})^2}{N}} &\neq \sqrt{\frac{(\hat{y}_i - y_i)^2}{N}} \\
\text{SD}(\hat{y}) &\neq \text{RMSE}(\hat{y}).
\end{aligned}
$$

\vspace{0.25in}

```{r regr_ex, echo = FALSE}
## set random seed for reprodicibility
set.seed(123)

## function to calculate RMSE
rmse <- function(pre, obs) {
  rmse <- sqrt(sum((pre - obs)^2)/length(pre))
  return(rmse)
}

## sample size
nn <- 20
## some values for the predictor x
xx <- sort(runif(nn, 0, 20))
## intercept
alpha <- 1
## slope
beta <- 0.5
## random errors (mean = 0, SD = 1)
epsilon <- rnorm(nn)
## observed data
yy <- alpha + beta*xx + epsilon

## fit a regression model
mm <- lm(yy ~ xx)
## get predicted values
yhat <- predict(mm)
```

# An example

Here is a simple example that shows how $\text{SD}(\hat{y})$ and $\text{RMSE}(\hat{y})$ are different. Consider a case where we had reason to believe that a variable $y$ was a function of another variable $x$. In effort to undercover the nature of their relationship, we collected 20 samples of both $y$ and $x$ (Figure 1).

```{r fig_1_data, echo = FALSE, fig.height=4.5, fig.width=4, fig.pos="placeHere", fig.align="center"}
par(mai = c(1.5,1,0.5,0.25), cex = 0.7)
## plot yhat and mean(yhat)
plot(xx, yy, pch = 19, las = 1,
     xlab = expression(italic(x)),
     ylab = expression(italic(y)))
```

`r fig_cap(cnt, caption = "Plot of some hypothetical data.", display = "full")` 

Based on the apparent relationship between $x$ and $y$, we might assume that each of the observed values $y_i$ is a linear combination of an intercept $\alpha$, the effect $\beta$ of a covariate $x_i$, and some random observation error $\epsilon_i$, such that

$$
y_i = \alpha + \beta x_i + \epsilon_i,
$$

and $\epsilon_i \sim \text{N}(0, \sigma)$. We could easily estimate the unknown parameters in this model ($\alpha$, $\beta$, $\sigma$), and then use the deterministic portion of the model to make predictions to compare with each of the observed values. Specifically, the predictions $(\hat{y}_i)$ would be given by a straight line, such that

$$
\hat{y}_i = \hat{\alpha} + \hat{\beta} x_i.
$$

We could then use these predictions to estimate both their SD and RMSE (Figure 2). The SD of $\hat{y}$ is ~`r round(sd(yhat), 3)`, but the RMSE is only ~`r round(rmse(yhat, yy), 3)`.

\vspace{0.2in}


```{r plot_regr_ex, echo = FALSE, fig.width = 8, fig.height = 4.5, fig.align="center"}
par(mfrow = c(1,2), mai = c(1.5,1,0.5,0.25), cex = 0.7)
## plot yhat and mean(yhat)
plot(xx, yy, pch = 16, las = 1, type = "n",
     xlab = expression(italic(x)),
     ylab = expression(italic(y)~~or~~m[hat(italic(y))]))
mtext("SD", side = 3, line = 1)
segments(x0 = xx, y0 = yhat, y1 = mean(yhat), col = "gray")
points(xx, yhat, pch = 21, bg = "white")
points(xx, rep(mean(yy), nn), pch = 19)
## plot observed and yhat
plot(xx, yy, pch = 16, las = 1, type = "n",
     xlab = expression(italic(x)),
     ylab = expression(italic(y)~~or~~hat(italic(y))))
mtext("RMSE", side = 3, line = 1)
segments(x0 = xx, y0 = yhat, y1 = yy, col = "gray")
points(xx, yy, pch = 19)
points(xx, yhat, pch = 21, bg = "white")
```

```{r, echo=FALSE}
cnt <- cnt + 1
```

`r fig_cap(cnt, caption = "Graphical examples of the difference between the SD of the predictions (left) and the RMSE of the predictions (right). For the SD, the comparison is based upon the differences between the predictions (open circles) and their mean (filled circles). For the RMSE, the comparison is based upon differences between the predictions (open circles) and the observed data (filled circles). In both cases, one would square the length of each of the vertical gray lines, sum them up, and divide by the number of them before finally taking the square root.", display = "full")` 

